---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.width = 6,
  fig.height = 4
)
```

# ggbulksurv

<!-- badges: start -->
<!-- badges: end -->

**What does this package do?**

`ggbulksurv` aims to simplify survival analysis for _C.elegans_ and _D.melanogaster_ by creating a user-friendly pipeline. Lifespan data for these organisms typically look like this, which is incompatible with current packages:

condition | day | dead | censored
--------  | --- | ---- | --------
WT        | 10  | 4    | 2
Drug1     | 10  | 9    | 0
...       | ... | ...  | ....

In particular, we highlight these two features: 

1. [Convert lifespan data for GraphPad PRISM](https://qhuitan.github.io/ggbulksurv/articles/convert-prism.html): Easily convert lifespan data into GraphPad PRISM compatible formats with the `pivot_prism()` function. 

2. [Easily customized survival curves in R](https://qhuitan.github.io/ggbulksurv/articles/customizing-plot.html): Use the `run_bulksurv()` command for a default analysis (plot + stats). Customizations: subsetting data, changing colors, changing p-value adjustment methods, etc. 

`ggbulksurv` is very much under **active development**, and any feedback and contributions are welcome through the [Issues](https://github.com/qhuitan/ggbulksurv/issues) page.

**Do I need this package?**

If you are conducting lifespan studies with _C.elegans_ and _D.melanogaster_ - probably. 

## Installation

Install the following dependencies from CRAN. If you use the `tidyverse`, these should already exist: 
``` r
install.packages(c("dplyr", "ggplot2", "janitor", "magrittr", "scales", "tidyr"))
```

Install two other dependencies from Bioconductor: 
``` r
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(c("survival", "survminer"))
```

You can install the development version of `ggbulksurv` from [GitHub](https://github.com/) with:

``` r
install.packages("devtools")
```

- For Mac users: If the above code gives an error, install Xcode. The easiest way to do this is via [Homebrew](https://brew.sh), a convenient package manager which automatically installs Xcode.

```r
devtools::install_github("qhuitan/ggbulksurv")
```

## Converting data for PRISM

```{r setup}
library(ggbulksurv) # load library
```


`pivot_prism()` converts bulk survival data into a GraphPad PRISM-compatible format.
Your input data file should be a `csv` with 4 columns: `condition`, `day`, `dead`, `censored`. Additional columns will be removed from the analysis. 

```{r}
data(sample_data)
dat <- sample_data # load fictional sample data
# Read in your data. Remove the # in front of the following line and replace the filename with your actual file
# dat <- read.csv("your-csv-file.csv")
```

```{r}
df_prism <- pivot_prism(dat)

head(df_prism) # A quick look
```

```{r, eval=FALSE}
# Export to csv
write.csv(df_prism, file = "lifespan_prism.csv")
```

## ggbulksurv (R)

### 1. Read in your data with `read.csv()`. 

Your data file should be a `csv` with 4 columns: `condition`, `day`, `dead`, `censored`. Additional columns (`sex`, `genotype`, `treatment` etc) are **allowed** and will be retained in this analysis.

```{r}
library(ggbulksurv)
```


Read in your `.csv` file with the following line of code: 

```{r}
data(sample_data)
dat <- sample_data # load example data

# Read in your data. Remove the # in front of the following line and replace the filename with your actual file
# dat <- read.csv("your-csv-file.csv")
```

### 2. Run the survival analysis: 

```{r survival}
# Plot a survival curve
p <- run_bulksurv(dat,
                  print_stats = TRUE # print stats. FALSE by default
                  )
```

Plotting a mortality curve: 

```{r mortality}
# Plot a mortality curve
p <- run_bulksurv(dat,
                  type = "mortality"
                  )
```

Custom formula: 

```{r custom-formula}
p <- run_bulksurv(dat, 
                  # Specify custom survival formula: condition + sex
                  formula = "Surv(day, status) ~ condition + sex", 
                  # Split colors by condition column
                  color = "condition", 
                  # Change linetype by sex column
                  linetype = "sex", 
                  palette = c("red", "blue", "forestgreen"))
```



