---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.width = 6,
  fig.height = 4
)
```

# ggbulksurv

<!-- badges: start -->
<!-- badges: end -->

The`ggbulksurv` package provides functions to facilitate bulk survival analysis, allowing users to input multiple observations per row.

**What is bulk survival analysis?** 

For some model organisms (eg _C. elegans_ and _D. melanogaster_), lifespan studies are usually performed in bulk. In bulk survival analysis, researchers place a certain number of organisms in a vial, and count the number of organisms that are dead/censored on any particular day. This produces a table that looks like this, with multiple observations per row:

condition | day | dead | censored
----      | --  | --   | ----
WT        | 10  | 2    | 0
WT        | 12  | 0    | 2
...

Unsurprisingly, survival analysis packages in R (eg `survival`, `survminer`) require each row to correspond to one individual. Wrangling the data manually is often a rather tedious task.

**What does this package do?**

`ggbulksurv` converts bulk survival data into individual observations per row, and plots a survival curve. Other functions are also available to plot mortality curves, customize colors, and to calculate relevant statistics such as median survival and log-rank tests.

`ggbulksurv` is very much under **active development**, and any feedback and contribution are welcome through the Issues page.

**What is R, and how do I get started?**

Download the most recent versions of R and RStudio for the appropriate OS using the links below:

- [R](https://cran.r-project.org)
- [RStudio](https://posit.co/download/rstudio-desktop/#download)

If you're new to R, I highly recommend this great resource by the Harvard Chan Bioinformatics Core. It provides a great overview of the RStudio interface, as well as R Projects that help keep your analysis organized. 

[https://hbctraining.github.io/Intro-to-R-flipped/lessons/01_introR-R-and-RStudio.html](https://hbctraining.github.io/Intro-to-R-flipped/lessons/01_introR-R-and-RStudio.html)

**Do I need this package?**

If you are conducting lifespan studies with _C.elegans_ and _D.melanogaster_ - probably. 


## Installation

Install the following dependencies from CRAN. If you use the `tidyverse`, these should already exist: 
``` r
install.packages(c("dplyr", "ggplot2", "janitor", "magrittr", "methods", "scales", "tidyr"))
```

Install two other dependencies from Bioconductor: 
``` r
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(c("survival", "survminer"))
```

You can install the development version of `ggbulksurv` from [GitHub](https://github.com/) with:

``` r
if (!require("devtools", quietly = TRUE))
    install.packages("devtools")
devtools::install_github("qhuitan/ggbulksurv")
```

## Getting started (quick)

### 1. Read in your data with `read.csv()`. 

Your data file should be a `csv` with 4 columns: `condition`, `day`, `dead`, `censored`. Additional columns will be removed from the analysis. 

```{r setup}
library(ggbulksurv)
library(survival)
library(survminer)
```


Read in your `.csv` file with the following line of code: 
```
dat <- read.csv("your-csv-file.csv")
```

For the purposes of this tutorial, I've created a ficitonal sample dataset, `sample_data` that we will use to illustrate the functions in this package. 

```{r}
data(sample_data)
dat <- sample_data # load sample data

head(dat)
```


### 2. Run the survival analysis: 

`ggbulksurv` can be run with the default settings using the `run_bulksurv()` command.

Plotting a survival curve: 

```{r survival}
# Plot a survival curve
p <- run_bulksurv(dat, 
                  sample_order = c("WT", "Drug1", "Drug2"), 
                  type = "survival")
```

Plotting a mortality curve: 

```{r mortality}
# Plot a mortality curve
p <- run_bulksurv(dat, 
                  sample_order = c("WT", "Drug1", "Drug2"), 
                  type = "mortality",
                  print_stats = FALSE # don't print stats
                  )
```

## Further customizations

### Changing the p-adjust method

If needed, we can change the p.adjust method. `run_bulksurv()` accepts the following corrections: "holm", "hochberg", "hommel", "bonferroni", "BH", "BY", "fdr", "none". Default: "BH"

```{r}
p <- run_bulksurv(dat, 
                  sample_order = c("WT", "Drug1", "Drug2"), 
                  type = "survival",
                  p_adjust_method = "bonferroni" # use bonferroni correction
                  )
```


### Changing colors

```{r surv_custom}
p <- run_bulksurv(dat, 
                  sample_order = c("WT", "Drug1", "Drug2"),
                  print_stats = FALSE,                   # don't print stats
                  palette = c("black", "red", "purple"), # Custom colors
                  legend.title = "",                     # Remove legend title
                  legend.position = c(0.9, 0.9),         # Position legend at top right
                  add.pval = TRUE                        # Add pvalue
             )
```

### Subsetting data

What if you're only interested in two conditions (eg WT vs Drug1)? 

```{r}
# Specify your conditions of itnerest here
conditions_of_interest <- c("WT", "Drug1")


dat_filt <- dat %>% 
  # Keep rows where condition is in conditions_of_interest
  dplyr::filter(condition %in% conditions_of_interest) 

# Plot
p_filt <- run_bulksurv(dat_filt, 
                  sample_order = c("WT", "Drug1"), 
                  type = "mortality",
                  print_stats = FALSE # don't print stats
                  )
```



## Other functionalities

### Interfacing with PRISM

GraphPad PRISM remains a favourite among biologists. To allow bulk survival data to be quickly converted to a PRISM-compatible format, use the `pivot_prism` function: 

```{r}
df_prism <- pivot_prism(dat, 
                        sample_order = c("WT", "Drug1", "Drug2"))

head(df_prism) # A quick look
```

```{r, eval=FALSE}
# Export to csv
write.csv(df_prism, file = "lifespan_prism.csv")
```



```{r}
sessionInfo()
```

