---
title: "Basic plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{basic-plot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(rmarkdown.html_vignette.check_title = FALSE)
```


## Load libraries

```{r setup}
library(ggbulksurv)
```

## Import csv

You should have 4 columns of data that look like this: 

condition | day | dead | censored
----      | --  | --   | ----
WT        | 10  | 2    | 0
WT        | 12  | 0    | 1
...       |...  | ...  | ...


```{r}
# This is a fictional dataset that is bundled with the package, and is used
# for the purposes of this tutorial
data(sample_data)
dat <- sample_data

head(dat) # see the first few rows
```


To import your own data, do the following: 

1. Export your Excel table as a `csv` file (eg: `your-csv-file.csv`). Note that all your conditions must be in __the same__ sheet. 
2. Create an R project in the __same folder__ as your `csv` file. 
3. Use the following line of code to import your `csv` file into R. 

```
dat <- read.csv("your-csv-file.csv")
head(dat) # see the first few rows
```

What conditions are present? 

```{r}
unique(dat$condition)
```

## Survival plot

`run_bulksurv()` plots a survival curve, and outputs median survival, log-rank test and pairwise log-rank test statistics: 

```{r}
p <- run_bulksurv(dat) # Default: survival curve
```

## Mortality plot

Use `type = "mortality"` for a mortality curve:

```{r}
p <- run_bulksurv(dat, 
                  print_stats = FALSE, 
                  type = "mortality")
```


