---
title: "Convert to PRISM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{convert-prism}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

GraphPad PRISM requires a very specific input format for lifespan analysis. In this tutorial, we show how the `pivot_prism()` function can be used to easily convert data for PRISM. 

## Setup

```{r setup}
# Load libraries
library(ggbulksurv)
```


```{r}
# Load data
data(sample_data)
dat <- sample_data # using the fictional dataset

# Load your data for your analysis
# dat <- read.csv("your-sample-data.csv")
```

What conditions are present? Use this to check for any accidental typos: 

```{r}
unique(dat$condition) 
```

## Pivot to PRISM
```{r}
df_prism <- pivot_prism(dat)

head(df_prism) # quick look
```


## Export
That looks great! Let's export it: 

```{r eval = FALSE}
write.csv(df_prism, file = "df_prism.csv")
```

You should see a `df_prism.csv` file appear in your folder. Copy and paste this into GraphPad PRISM, and run a survival analysis in PRISM. 


