---
title: "Customizing Plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{customizing-plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup}
library(ggbulksurv)
```

This vignette teaches you how to further customize your survival plots. 

We'll start by discussing how to change the colors of the plot, using either custom colors, RColorBrewer palettes or journal colors. 

We'll then discuss how to customize the legend positions and titles. 

Lastly, we finish off by adding the 50% median survival line. 


## Color palettes

### Brewer palettes

[RColorBrewer](https://r-graph-gallery.com/38-rcolorbrewers-palettes.html) provides a list of useful, aesthetically pleasing palettes that are frequently used in R. To use these palettes, change the `palette` parameter: 

```{r}
# Using the Set1 brewer palette
p <- run_bulksurv(sample_data, 
             sample_order = c("WT", "Drug1", "Drug2"),
             print_stats = FALSE,
             palette = "Set1" # Custom brewer palette
             )
```


### Journal palettes

[ggsci](https://nanx.me/ggsci/index.html) provides a collection of palettes inspired by scientific publishing groups. These colors can also be used in our plots: 

```{r}
# Using the nature palette
p <- run_bulksurv(sample_data, 
             sample_order = c("WT", "Drug1", "Drug2"),
             print_stats = FALSE,
             palette = "npg" # Nature palette
             )
```


### Specific colors

We can also specify colors, either by name ("black", "blue"), or by hexadecimal value ("#80080"): 

```{r}
# Using manual colors
p <- run_bulksurv(sample_data, sample_order = c("WT", "Drug1", "Drug2"),
             print_stats = FALSE,
             palette = c("black", "red", "#800080") # Custom colors
             )
```

## Legends

### Removing legend title

```{r}
p <- run_bulksurv(sample_data, sample_order = c("WT", "Drug1", "Drug2"),
             print_stats = FALSE,
             legend.title = "" # Remove the legend title
             )
```


### Specifying legend position
Specifying a position (`right`, `left`, `bottom` or `top`) with `legend.position`:

```{r}
p <- run_bulksurv(sample_data, 
                  sample_order = c("WT", "Drug1", "Drug2"),
             print_stats = FALSE,
             legend.position = "bottom")
```

### Specific legend coordinates

Specific coordinates can also be used. For bottom left, use `legend.position = c(0,0)`; for top right, use `legend.position = c(1,1)`.

```{r}
p <- run_bulksurv(sample_data, 
                  sample_order = c("WT", "Drug1", "Drug2"),
                  print_stats = FALSE,
                  legend.position = c(0.9, 0.9))
```


## Axis customizations

### Change the x-axis breaks

```{r}
p <- run_bulksurv(sample_data, sample_order = c("WT", "Drug1", "Drug2"),
             print_stats = FALSE
             )
```

## Annotations

### Adding pvalue and 50% median survival line

```{r}
p <- run_bulksurv(sample_data, sample_order = c("WT", "Drug1", "Drug2"),
             print_stats = FALSE,
             add.median.survival = TRUE, # Add median survival
             add.pval = TRUE # Add pvalue
             )
```

### Add confidence interval

```{r custom-addci}
p <- run_bulksurv(sample_data, sample_order = c("WT", "Drug1", "Drug2"),
             print_stats = FALSE,
             add.conf.int = TRUE
             )
```



### Putting it all together

```{r}
p <- run_bulksurv(sample_data, 
                  sample_order = c("WT", "Drug1", "Drug2"),
                  print_stats = FALSE,
                  add.median.surv = TRUE, 
                  palette = c("black", "red", "purple"), # Custom colors
                  legend.title = "",                     # Remove legend title
                  legend.position = c(0.9, 0.9),         # Position legend at top right
                  add.pval = TRUE                       # Add pvalue
             )

p
```

