---
title: "Customizing Plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{customizing-plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.height = 4,
  fig.width = 6
)

options(rmarkdown.html_vignette.check_title = FALSE)
```


`ggbulksurv` offers several customization options:

- Subsetting data
- Colors
- Legend title and position
- p-value annotations

This tutorial will go through several most commonly used customizations. To see all possible options, type `?run_bulksurv()` in the console. 


```{r setup}
library(ggbulksurv)
```


```{r}
# Load sample_data (the fictional dataset)
data(sample_data)
dat <- sample_data
```

## The default

For comparison, here's the default plot: 

```{r}
p <- run_bulksurv(dat,
                  print_stats = FALSE)
```


## A customized survival curve

Here is an example of a highly customized survival curve: 

```{r}
p <- run_bulksurv(dat, 
                  sample_order = c("WT", "Drug1", "Drug2"),  # Custom order
                  print_stats = FALSE,                   # Don't print stats
                  palette = c("black", "red", "purple"), # Custom colors
                  legend.title = "",                     # Remove legend title ("Condition")
                  legend.position = c(0.9, 0.9),         # Position legend at top right
                  add.pval = TRUE                       # Add pvalue
             )
```


## Changing colors

### Brewer palettes

[RColorBrewer](https://r-graph-gallery.com/38-rcolorbrewers-palettes.html) provides a list of useful, aesthetically pleasing palettes that are frequently used in R. To use these palettes, change the `palette` parameter: 

```{r}
# Using the Set1 brewer palette
p <- run_bulksurv(sample_data, 
             sample_order = c("WT", "Drug1", "Drug2"),
             print_stats = FALSE,
             palette = "Set1" # Custom brewer palette
             )
```

### Journal palettes

[ggsci](https://nanx.me/ggsci/index.html) provides a collection of palettes inspired by scientific publishing groups. These colors can also be used in our plots: 

```{r}
# Using the nature palette
p <- run_bulksurv(sample_data, 
             sample_order = c("WT", "Drug1", "Drug2"),
             print_stats = FALSE,
             palette = "npg" # Nature palette
             )
```

### Custom colors (name, hexadecimal)

We can also specify colors, either by name ("black", "blue"), or by hexadecimal value ("#80080"): 

```{r}
# Using manual colors
p <- run_bulksurv(sample_data, sample_order = c("WT", "Drug1", "Drug2"),
             print_stats = FALSE,
             palette = c("black", "red", "#800080") # Custom colors
             )
```

## Legends

### Remove legend title

```{r}
p <- run_bulksurv(sample_data, 
                  sample_order = c("WT", "Drug1", "Drug2"),
             print_stats = FALSE,
             legend.title = "" # Remove the legend title
             )
```

### Specify legend position

Specifying a position (`right`, `left`, `bottom` or `top`) with `legend.position`:

```{r}
p <- run_bulksurv(sample_data, 
                  sample_order = c("WT", "Drug1", "Drug2"),
             print_stats = FALSE,
             legend.position = "bottom" # Set legend at bottom
             )
```

Note that `run_bulksurv` returns the plot as a `ggplot` object, so you can further customize it if required: 

```{r fig.height=4, fig.width = 6}
p + 
  ggplot2::theme(legend.position = "bottom") # Change legend position to be at bottom
```


### Specify legend coordinates

Specific coordinates can also be used. For bottom left, use `legend.position = c(0,0)`; for top right, use `legend.position = c(1,1)`.

```{r}
p <- run_bulksurv(sample_data, 
                  sample_order = c("WT", "Drug1", "Drug2"),
                  print_stats = FALSE,
                  legend.position = c(0.9, 0.9))
```

## P-value and median survival line

To add the median survival line and pvalues, pass the following parameters within `run_bulksurv()`:

- median survival line: `add.median.survival = TRUE"`
- pvalue annotation: `add.pval = TRUE`

To change the p-value adjustment method, use `p_adjust_method`. 

```{r}
p <- run_bulksurv(sample_data, 
                  sample_order = c("WT", "Drug1", "Drug2"),
             print_stats = FALSE,
             add.median.survival = TRUE, # Add median survival
             p_adjust_method = "bonferroni", # Change padj method
             add.pval = TRUE             # Add pvalue
             )
```

## Adding confidence intervals

```{r custom-addci}
p <- run_bulksurv(sample_data, sample_order = c("WT", "Drug1", "Drug2"),
             print_stats = FALSE,
             add.conf.int = TRUE  # Add confidence interval
             )
```

## Returning data

By default, `run_bulksurv()` returns a `ggplot2` object for easy plot customization. 

If you want to return the pivoted data, summary statistics and the plot object, use `returnData = TRUE`. This will return all data and plots as a `list` object. 

```{r}
df_survival <- run_bulksurv(dat, returnData = TRUE)

names(df_survival)
```

Accessing the individual elements: 
```{r}
head(df_survival$data) # pivoted data
```


```{r}
head(df_survival$median_survival) # median survival
```


```{r}
head(df_survival$logrank) # logrank test
```


```{r}
head(df_survival$pairwise) # pairwise logrank with p-value correction
```


