---
title: "ggbulksurv v0.2.0"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{quick-start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(rmarkdown.html_vignette.check_title = FALSE)
```

**What does this package do?**

`ggbulksurv` aims to simplify bulk survival analysis by creating a default pipeline for _C.elegans_ and _D.melanogaster_. In particular, we highlight these two features: 

1. `pivot_prism()` : Easily converts lifespan data into GraphPad PRISM compatible formats
2. `run_bulksurv()` : A one-stop command to plot default survival curves and perform statistical analysis. 

`ggbulksurv` is very much under **active development**, and any feedback and contributions are welcome through the Issues page.

**Do I need this package?**

If you are conducting lifespan studies with _C.elegans_ and _D.melanogaster_ - probably. 

**What is R, and how do I get started?**

For complete beginners to R, see `vignette("getting-started-with-r")` to complete setup. 
...

## Setup

```{r setup}
## Load libraries
library(ggbulksurv)
```

### Read in the `csv` file

If you're using Excel, you should have 4 columns of data that look like this: 

condition | day | dead | censored
----      | --  | --   | ----
WT        | 10  | 2    | 0
WT        | 12  | 0    | 1


Export your Excel table as a `csv` file. Create an R project in the same folder as your `csv` file, and use the following line of code to import your `csv` file into R. 

```
dat <- read.csv("your-csv-file.csv")
```

```{r}
# Load sample_data (the fictional dataset)
data(sample_data)

# Save this as variable dat so it's easier for us to call it
dat <- sample_data
```

## Convert to PRISM

Use the `pivot_prism()` function to convert data for GraphPad PRISM. 

```{r}
df_prism <- pivot_prism(dat) # pivot the data

head(df_prism) # quick preview
```

```{r eval = FALSE}
write.csv(df_prism, file = "df_prism.csv", row.names = FALSE) # export
```

You can import `df_prism.csv` into GraphPad PRISM for further analysis. 

## run_bulksurv

We run a default analysis using the `run_bulksurv` function: 

```{r}
p <- run_bulksurv(dat)
```



