---
title: "Quick Start"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{quick-start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(rmarkdown.html_vignette.check_title = FALSE)
```

## Setup

### Load libraries
```{r setup}
library(ggbulksurv)
```

### Read in the `csv` file

If you're using Excel, you should have 4 columns of data that look like this: 

condition | day | dead | censored
----      | --  | --   | ----
WT        | 10  | 2    | 0
WT        | 12  | 0    | 1


Export your Excel table as a `csv` file. Create an R project in the same folder as your `csv` file, and use the following line of code to import your `csv` file into R. 

```
dat <- read.csv("your-csv-file.csv")
```

To illustrate the functions available in this package, I've created a small fictional dataset. If you would like to follow this tutorial with this sample dataset, load it with this code: 

```{r}
# Load sample_data (the fictional dataset)
data(sample_data)

# Save this as variable dat so it's easier for us to call it
dat <- sample_data
```


Let's take a quick look at `dat`: 

```{r}
head(dat)
```

As expected, `dat` has 4 columns: `condition`, `day`, `dead`, `censored`. 

What conditions are present? 

```{r}
unique(dat$condition)
```


## run_bulksurv

### Default analysis

We run a default analysis using the `run_bulksurv` function: 

```{r}
p <- run_bulksurv(dat, 
                  sample_order = c("WT", "Drug1", "Drug2"), 
                  type = "survival")
```

The plot dimensions don't look too great. To fix it, we specify the dimensions (fig.height = 4, fig.width = 6)


```{r fig.height=4, fig.width = 6}
p <- run_bulksurv(dat, 
                  sample_order = c("WT", "Drug1", "Drug2"), 
                  type = "survival")
```


Note that `run_bulksurv` returns the plot as a `ggplot` object, so you can further customize it if required: 

```{r fig.height=4, fig.width = 6}
p + 
  ggplot2::theme(legend.position = "bottom") # Change legend position to be at bottom
```

### Only print the plot

By default,`run_bulksurv` prints both summary statistics and the plot. You can turn it on or off using the `print_stats` and `print_plot` parameters, respectively. 

```{r fig.height=4, fig.width = 6}
# Turning off summary statistics

p1 <- run_bulksurv(sample_data, sample_order = c("WT", "Drug1", "Drug2"), 
             print_stats = FALSE, # don't print stats
             print_plot = TRUE # print plot
             )
```

### Only print statistics

```{r}
p2 <- run_bulksurv(dat, 
                   sample_order = c("WT", "Drug1", "Drug2"), 
                   print_stats = TRUE,
                   print_plot = FALSE # don't print plot
             )
```

## Conclusion

In this vignette, we explored the default `run_bulksurv` commands. For further customizations, explore these vignettes: 

- [Intermediate customizations](`vignette("customizing-plots.Rmd")`). Customize colors, axis and themes using 'run_bulksurv()'. 
- [Advanced customizations](`vignette("advanced-customizations.Rmd")`). A deep dive behind the three functions that make up run_bulksurv(), as well as interfacing with 'survminer' and 'survival' packages. 


