---
title: "Interfacing with other survival packages"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{advanced-customizations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup}
library(ggbulksurv)
library(survival)
library(survminer)
```


Two of the most popular packages for survival analysis are [survival](https://cran.r-project.org/web/packages/survival/index.html) and [survminer](https://cran.r-project.org/web/packages/survminer/index.html). This vignette will demonstrate how to use `ggbulksurv`'s output as input to other survival packages. 

We first start by documenting the three functions that `run_bulksurv()` wraps around. At each stage, we demonstrate how this can be used as input to other packages for the advanced user. 

## Behind the scenes

The `run_bulksurv()` command makes several choices for the user, with the caveat that these assumptions tend to hold under most conditions. Users who desire total control over the process should read this section for a more detailed walkthrough.

`run_bulksurv()` is a wrapper around the 3 following functions: 

1. `get_indiv_surv`: Converts a bulk survival table into an individual survival table.
2. `fit_surv`: Fits a survival object from `survival::survfit()`.
3. `plot_surv`: Plots a survival curve using `survminer::ggsurvplot()`.

This section takes you through the 3 steps that `run_bulksurv()` wraps around. 

## 1. Get individual survivals

`get_indiv_surv` converts a table of bulk survival data into individual survivals. Each row now represents an individual. `day` represents the number of days lived, while `status` is either 0 (censored) or 1 (dead). 


```{r}
df_isurv <- get_indiv_surv(sample_data,
                           sample_order = c("WT", "Drug1", "Drug2"))

head(df_isurv)
```

We now have a `tibble` with one individual per row. 

At this point, this data can be used as input to other packages, such as  `survminer()` and `survival`. 

## 2. Fit the survival object

`fit_surv` creates a `survfit` object for `day` and `status`, then fits a survival curve by `condition`. It takes in a `data.frame` with one individual per row, and creates a `survfit` object using the `survival::survfit()` function. 

```{r}
# Fit survival object
df_fit <- fit_surv(df_isurv)

head(df_fit)
```

Alternatively, you can fit this manually using the `survival` package: 

```{r}
df_isurv <- survival::survfit(Surv(day, status) ~ condition,
                                data = df_isurv)
```

Drug1 has the shortest median lifespan of 4.5 days, with a 95% confidence interval (95% CI) of 4 to 6 days. In contrast, Drug2 has the longest median lifespan of 33 days, with a 95% CI of 32 to 35 days.


## 3. Plot the survival object

### Using plot_surv()

`plot_surv` is a wrapper around [survminer::ggsurvplot()]. As such, all additional arguments passed to `plot_surv` will be passed to `survminer::ggsurvplot()`. Some useful examples are illustrated below: 

```{r surv_detail}
df_isurv <- get_indiv_surv(sample_data,
                           sample_order = c("WT", "Drug1", "Drug2"))

plot_surv(fit = df_fit, 
          type = "survival",
          data = df_isurv)
```

#### Changing legend positions

Specifying a position (`right`, `left`, `bottom` or `top`) with `legend.position`:

```{r custom-legright}
plot_surv(df_fit, 
          type = "survival", 
          data = df_isurv,
          sample_order = c("WT", "Drug1", "Drug2"),
          print_stats = FALSE,
          legend.position = "right" # Change legend position
          )
```

Specific coordinates can also be used. For bottom left, use `legend.position = c(0,0)`; for top right, use `legend.position = c(1,1)`.

```{r custom-legcoord}
plot_surv(df_fit, 
          type = "survival", 
          sample_order = c("WT", "Drug1", "Drug2"),
          print_stats = FALSE,
          data = df_isurv,
          legend.position = c(0,0) # Specify coordinates
          )
```

It doesn't fit - in fact, the legend is partially obscuring our x and y axes. 

We can simplify the legend by removing the legend title and the trailing `condition=` symbol: 

```{r custom-rmlegtitle}
sample_order = c("WT", "Drug1", "Drug2")

plot_surv(df_fit, 
          type = "survival", 
          sample_order = sample_order,
          data = df_isurv,
          print_stats = FALSE,
          legend.position = c(0.09,0.2), # Specify coordinates
          legend.title = "", # Remove legend title
          legend.labs = sample_order # Remove `condition=`
          )
```

#### Adding confidence intervals and median survival

```{r custom-addci}
plot_surv(df_fit, 
          data = df_isurv,
          type = "survival",
          add.conf.int = TRUE, # Add 95% confidence interval
          legend.position = "bottom")
```

```{r custom-addmedsurv}
plot_surv(df_fit, 
          data = df_isurv,
          type = "survival",
          add.median.survival = TRUE, # Add median survival line
          legend.position = "bottom")
```

#### Customizing x-axis day breaks

```{r custom-xbreak}
plot_surv(df_fit, 
          data = df_isurv,
          type = "survival",
          legend.position = "bottom",
          break.x.by = 5 # Breaks every 5 days
          )
```

#### Putting it all together: 

```{r custom-1}
sample_order = c("WT", "Drug1", "Drug2")

plot_surv(df_fit, 
          data = df_isurv,
          type = "survival",
          sample_order = sample_order,
          legend.labs = sample_order,            # Rename legend
          print_stats = FALSE,                   # don't print stats
          palette = c("black", "red", "purple"), # Custom colors
          legend.title = "",                     # Remove legend title
          legend.position = c(0.9, 0.9),         # Position legend at top right
          add.pval = TRUE,                        # Add pvalue
          add.median.survival = TRUE             # Add 50% median line
             )
```


### Using survminer::ggsurvplot()

Alternatively, you can construct your own plots using `survminer::ggsurvplot`: 

```{r ggsurvplot, fig.height=6}
survminer::ggsurvplot(fit = df_fit,
                      data = df_isurv,
                      ncensor.plot = TRUE)
```

## 4. Print summary statistics

### Using the `summary_stats` function: 

Calculates median survival, logrank test and pairwise logrank test. 

```{r}
stats <- summary_stats(df_isurv, 
                       type = "all")

stats
```

This returns a `list` object, which allows us to individually pull out the values of interest. For example: 

#### The logrank test: 

```{r}
stats$logrank
```

```{r}
# chisq value of logrank test
stats$logrank$chisq
```


```{r}
# pvalue of logrank test
stats$logrank$pvalue
```

#### Pairwise log-rank test: 
```{r}
stats$pairwise
```

```{r}
# p-values
stats$pairwise$p.value
```

To use a different p-value adjustment method, specify the p_adjust_method as follows:

```{r}
stats_2 <- summary_stats(df_isurv, 
                       type = "all", 
                       p_adjust_method = "bonferroni")

stats_2
```

### Using the `survminer` package directly

```{r}
test = survminer::pairwise_survdiff(
  formula = survival::Surv(day, status) ~ condition, # Fit a survival object
  data = df_isurv,
  p.adjust.method = "bonferroni" # Use bonferroni correction
  )

test
```

```{r}
# obtain specific p-values
test$p.value
```

