---
title: "Interfacing with other survival packages"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{advanced-customizations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(rmarkdown.html_vignette.check_title = FALSE)
```


Two of the most popular packages for survival analysis are [survival](https://cran.r-project.org/web/packages/survival/index.html) and [survminer](https://cran.r-project.org/web/packages/survminer/index.html). This vignette will demonstrate how to use `ggbulksurv`'s output as input to other survival packages. 

We first start by documenting the three functions that `run_bulksurv()` wraps around. At each stage, we demonstrate how this can be used as input to other packages for the advanced user.

```{r setup}
library(ggbulksurv)
library(survival)
library(survminer)
```

 

## Behind the scenes

The `run_bulksurv()` command makes several choices for the user, with the caveat that these assumptions tend to hold under most conditions. Users who desire total control over the process should read this section for a more detailed walkthrough.

`run_bulksurv()` is a wrapper around the 3 following functions: 

1. `get_indiv_surv`: Converts a bulk survival table into an individual survival table.
2. `fit_surv`: Fits a survival object from `survival::survfit()`.
3. `plot_surv`: Plots a survival curve using `survminer::ggsurvplot()`.

This section takes you through the 3 steps that `run_bulksurv()` wraps around. 

```{r}
# Load lifespan data
data(sample_data)
dat <- sample_data

# To read in your own data, remove the # from the following line of code:
# dat <- read_csv("your-lifespan-data.csv")
```


## 1. Get individual survivals

`get_indiv_surv` converts a table of bulk survival data into individual survivals. Each row now represents an individual. `day` represents the number of days lived, while `status` is either 0 (censored) or 1 (dead). 


```{r}
# Convert to one individual per row
df_isurv <- get_indiv_surv(dat,
                           sample_order = c("WT", "Drug1", "Drug2"))

head(df_isurv)
```

We now have a `tibble` with one individual per row. 

At this point, this data can be used as input to other packages, such as `survminer()` and `survival`. 

## 2. Fit the survival object

### fit_surv()

`fit_surv` creates a `survfit` object for `day` and `status`, then fits a survival curve by `condition`. It takes in a `data.frame` with one individual per row, and creates a `survfit` object using the `survival::survfit()` function. 

```{r}
# Fit survival object
df_fit <- fit_surv(df_isurv)

head(df_fit)
```
Interpretation: Drug1 has the shortest median lifespan of 4.5 days, with a 95% confidence interval (95% CI) of 4 to 6 days. In contrast, Drug2 has the longest median lifespan of 33 days, with a 95% CI of 32 to 35 days.

### survival package

Alternatively, you can fit this manually using the `survival` package: 

```{r}
# Fit survival object
df_fit <- survival::survfit(Surv(day, status) ~ condition,
                                data = df_isurv)

head(df_fit)
```

## 3. Plot the survival object

### Using plot_surv()

`plot_surv` is a wrapper around [survminer::ggsurvplot()]. As such, all additional arguments passed to `plot_surv` will be passed to `survminer::ggsurvplot()`. 


```{r surv_detail}
# Using ggbulksurv
df_isurv <- get_indiv_surv(dat,
                           sample_order = c("WT", "Drug1", "Drug2"))

df_fit <- fit_surv(df_isurv)

plot_surv(fit = df_fit, 
          type = "survival",
          data = df_isurv)
```


### Using survminer::ggsurvplot()

Alternatively, you can construct your own plots using `survminer::ggsurvplot`: 

```{r ggsurvplot, fig.height=6}
df_isurv <- get_indiv_surv(dat,
                           sample_order = c("WT", "Drug1", "Drug2"))

df_fit <- fit_surv(df_isurv)

survminer::ggsurvplot(fit = df_fit,
                      data = df_isurv,
                      ncensor.plot = TRUE # add censor plot
                      )
```


## 4. Returning data

`ggbulksurv` also provides functions to return data if you wish to use it in a different pipeline. Use the `returnData = TRUE` parameter to return everything as a list. 

```{r}
df_survival <- run_bulksurv(dat, returnData = TRUE)

class(df_survival)
names(df_survival)
```

```{r}
head(df_survival)
```

