---
title: "subset-and-reorder"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{subset-and-reorder}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ggbulksurv)
```


# Single variable (default)

For most survival analysis, a simple, one-variable formula is used. 
Call: `Surv(day, status) ~ condition`

```{r}
p <- run_bulksurv(sample_data)
```

## Subsetting one-variable formulae

To compare fewer conditions, change the `sample_order` parameter. 

Here, we subset AND invert the order: 

```{r}
p <- run_bulksurv(sample_data, 
                  sample_order = c("Drug2", "WT"), # Compare WT vs Drug2, invert order
                  print_stats = FALSE)
```

Note: `sample_order` only works for reordering the `condition` column. If you have a more complex design, an error message will be shown. 

# More than 1 variable (complex)


What happens when we want to model the effect of condition (drug) and another variable, such as sex? 

Let's take a look at `sample_data`" 

```{r}
head(sample_data)
```


## Fitting a custom formula

Use the `formula` parameter to specify a custom survival formula. 

```{r}
p <- run_bulksurv(sample_data, 
                  formula = "Surv(day, status) ~ condition + sex")
```

## Customizing colors and linetypes

We can use the `palette` parameter to specify custom colors: 

```{r}
p <- run_bulksurv(sample_data, 
                  formula = "Surv(day, status) ~ condition + sex", 
                  # Specify colors
                  palette = c("lightgreen", "forestgreen", 
                            "lightblue", "blue", 
                            "pink", "firebrick"))
```


We can further customize colors by groups in the data. Here, we color by `condition` and split linetypes by `sex`. 

```{r}
p <- run_bulksurv(sample_data, 
                  formula = "Surv(day, status) ~ condition + sex", 
                  color = "condition", 
                  linetype = "sex", 
                  palette = c("red", "blue", "forestgreen"))
```







